---
title: "Basic Visualisation in R"
subtitle: "Data Visualisation and Analytics"
author: "Anastasios Panagiotelis and Lauren Kennedy"
date: "Lecture 3"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"css/mtheme.css","css/mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      
---

class: inverse, center, middle

# The Grammar of Graphics

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(DT)
require(PoEdata)
require(kableExtra)
require(gridExtra)
require(ggmosaic)
```

---

# The grammar of graphics

- At first using `ggplot2` can seem too complicated.
- Once mastered it can be used to very easily create detailed plots.
- It is built on the ideas of *Grammar of Graphics* a text by Leland Wilkinson.
- The objective is to find an abstract set of rules for creating almost any graphic.

---

# Data

- The starting point for all visualisation is a dataset.
--

- In these slides, we will consider the datasets `diamonds`,  `mpg` and `economics` which come built in with the `ggplot2` package.
--

- Later on we learn how to read in data.
--

- The `diamonds` data contains data on the price, size and quality of over 50000 diamonds.

---

# Aes and Geom

- Think of an *aesthetic* (or aes) as a way of perceiving a variable:
  + Position on x or y axis
  + Color
  + Size
- Think of a *geometry* (or geom) as a way of representing a variable:
  + Points
  + Lines
- `ggplot` maps aesthetics to geometries  

---
class: inverse, center, middle

#Histogram

---
# Histogram

- Consider a histogram of the variable price
--

- In a histogram, values of the variable we are interested in lie along the horizontal (x) axis.
--

- The histogram creates bins then counts the number of observations in each bin.
--

- To get started type

```{r, echo=T, eval=F}
ggplot(data = diamonds,mapping = aes(x=price))
```

---

# What do we see?

```{r, echo=F, eval=T}
ggplot(data = diamonds,mapping = aes(x=price))
```

---

# What do we see?

- We do have an x axis with a label price and some values.
--

- Otherwise we see nothing.
--

- We need to **add** a geometry to the plot.
--

- We do this with the `geom_histogram` function.
```{r, echo=T, eval=F}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram()
```

---

# What do we see?

```{r, echo=F, eval=T,message=FALSE}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram()
```

---

# Modification

- Suppose want to use a different number of bins or change the color of the bins?
--

- These are not features of the data or the aes
--

- These are features of the *geom*.  
--

- So these are controlled by arguments in the `geom_histogram` function.

---

# Change bins

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(bins = 5)
```

---

# Change boundary

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(bins =  5, boundary=0)
```
---

# Change binwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500)
```

---

# Change color

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,fill = 'red')
```

---

# Change border color

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,color='white',
                 fill = 'blue')
```

---
class:inverse, middle center

# An aside on colour

---

# Customise colour

- Many colours come built in to R.
--

- In some cases you may wish to select your own color.
--

- Customising colour requires appreciating how a computer understands color.
--

- We will do this by looking at RGB hex codes.
--

- Using this system, to a computer *#ff0000* is red.

---

# The RGB system

- One color model used by computers encodes every colour by the amount of *red*, *green* and *blue* light mixed to make that colour.
--

- This is called the RGB color model.
--

- A value between 0 and 255  indicates the strength of red, green and blue.
--

- These values between 0 and 255 are represented in two hexadecimal digits.

---

# Hexadecimal

- In hexadecimal: 
  - **a** is ten,
  - **b** is eleven,
  - **c** is twelve... 
  - **f** is fifteen.
- Take the first digit and multiply by 16 and add the second digit
- Hexadecimal is used since each digit corresponds to 4 bits in  computer memory.
---

#Examples

- **10** in hexadecimal is $1\times 16+0=16$ in decimal
- **1a** in hexadecimal is $1\times 16+10=26$ in decimal
- **2b** in hexadecimal is $2\times 16+11=42$ in decimal
- What is e4 in decimal?
---

# Color picker 

- One online tool to find the hex code of a color is <a href ="https://www.w3schools.com/colors/colors_picker.asp">here</a>.
--

- Suppose we want to the histogram to be <font color=#b35900> this brown</font> color.
--

- The hex code is #b35900 which is 179/256 red, 89/256 green and no blue.
--

- This can be provided as a string, to the `fill` or `color` argument of `geom_histogram`.

---

# Brown histogram

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,color='white',
                 fill = '#b35900')
```
---

# Finding hex codes

- It is useful to know hex codes since at times you may want to match colors for a specific purpose.
--

- For instance you may want the colors to match the brand colors of a client.
--

- For example a simple online search tells us that Coca Color red is <font color="#f40000">#f40000</font>
--

- The green color worn by NBA team the Milwaukee Bucks is <font color="#00471b">#00471b</font>.

---
# Histograms (Bucks Colors)

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(fill='#00471b',color='#eee1c6')
```
---

# An exercise

- Find the hex codes for a color(s) associated with:
  - A brand you like, or
  - A sports team you like, or
  - Your country's flag,
  - Anything else
- Construct a histogram of the variable *carat* with these colors.

---
class: inverse, middle, center

# Density plot

---

# Density

- For a smoother version of a histogram we can use a different geom called `geom_density`.
--

- This in fact computes a kernel density estimate of the variable.
--

- The level of smoothness is controlled by a *bandwidth* parameter
--

- All the computation is done by `ggplot2`.

---

# Density plot

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density()
```

---
# Density plot (thicker)

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(size=3)
```

---
# How is density calculated?

- The kernel density estimate is a popular *nonparametric* technique that estimates a density as

$$\hat{f}(x)=\frac{1}{n}\sum\limits_{i=1}^{n}K_h(x-x_i)$$

- Here, $K_h()$ is a kernel function that depends on a bandwidth $h$.

---

# Uniform kernel

- The simplest kernel function is the uniform kernel
  + $K_h(u)=1/h$ if $|u|<h$
  + $K_h(u)=0$ otherwise
--

- At a point $x$, the estimated density is proportional to the number of points that are *close* to $x$.
--

- By *close*, we mean within $h$ units of $x$.

---
# Extremes

- If the bandwidth gets extremely large then for any $x$, all sample points are considered close.
--

- The formula for the kernel density becomes a flat line.
--

- If the bandwidth gets extremely small then for any $x$ we choose, the density is just the number of points in the sample equal to $x$.
--

- The kernel density is made up of spikes at the sample points.

---
# Defaults

- By default, `geom_density` 
  - Uses a Gaussian kernel 
  - Selects the bandwidth using *Silverman's rule of thumb*
--

- The same principles apply:
  - Large bandwidth leads to more smoothness
  - Small bandwidth leads to more bumpiness

---

# Density plot: Low bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=100)
```

---

# Density plot: High bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=2000)
```
---
# Density plot: Low bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=0.0001)
```

---

# Density plot: High bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=80000)
```
---

# Summary

- With both histograms and density plots
  - If the bin width or bandwidth is too small the plot may look bumpy.  This can exaggerate features that are not significant.
  - If the bin width or bandwidth is too large the plot may smooth over important features like local modes.
- Always try a few different values of bin width or bandwidth.

---
class: inverse, middle, center

# Finding outliers

---

# Outliers

- Histograms and density plots give a good idea of shape and local modes.
--

- Sometimes they can obscure outliers.
--

- For finding outliers a rug plot can be useful
--

- For finding outliers while still getting a good idea of skew, boxplots can be useful.
--

- We can investigate using the variable *carat*

---

# Carat: Histogram

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=carat))+
  geom_histogram()
```

---

# Carat: Rug plot

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=carat))+
  geom_rug()
```

---

# Box plot

- The box plot summarises 5 numbers
  - Median
  - First quartile $Q1$
  - Third quartile $Q3$
  - Upper Fence $U = Q3+1.5\times(Q3-Q1)$
  - Lower Fence $L = Q1-1.5\times(Q3-Q1)$
- Anything lying outside the fences represented as dots.
- When no points lie outside the fence, the fence is set to the maximum or minimum.

---

# Carat: Boxplot

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(y=carat))+
  geom_boxplot()
```

---

# Change of aesthetic

- Notice that the aesthetic changed!
--

- In the boxplot, the value of the variable is represented by the vertical (or y axis).
--

- We can change the definition of the upper and lower fence by passing the `coef` argument to `geom_boxplot`.  
--

- This changes the 1.5 used in calculating the fence to whatever you specify

---

# Changing fences

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(y=carat))+
  geom_boxplot(coef=4)
```


---


# Notches

- Notches can be added to a boxplot
--

- These are set to $\frac{1.58\times(Q3-Q1)}{\sqrt{n}}$
--

- This roughly gives a 95% confidence interval for the median.
--

- We will use a smaller dataset on the mileage of cars for this example to clearly illustrate the notches.

---

# Notches

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = mpg,mapping = aes(y=cty))+
  geom_boxplot(notch = T)
```

---
class: inverse, middle, center

# One Non-Metric Variable

---

# Nominal v Ordinal

- Non-metric variables are made up of **nominal** and **ordinal** variables.
--

- Nominal variables have no ordering in the categories of data:
  - Manufacturer of car (Audi, Toyota, etc).
--

- Ordinal variables do have an ordering in the categories:
  - Quality of diamonds (Fair, Good, etc).
  
---

# Non-metric variables in R

- Non-metric variables can be stored in R as
  - Character variables (nominal data)
  - Factors (nominal data)
  - Ordered factors (ordinal data)
- You can check with the `str` function

---

# Diamonds data

```{r, echo=T}
str(diamonds)
```

---

# Mpg data

```{r, echo=T}
str(mpg)
```

---

#Bar plot

- A common plot for non-metric data is the bar plot for the frequency of observations for each level of the factor.
--

- The height of each bar indicates the number of observations in a particular category.
--

- This can be done using `geom_bar`

---

#Bar plot

```{r, echo=T, fig.height=5}
ggplot(data = diamonds, mapping = aes(x=cut))+
  geom_bar()
```

---

#Bar plot

```{r, echo=T, fig.height=5}
ggplot(data = mpg, mapping = aes(x=manufacturer))+
  geom_bar()
```

---


class: inverse, center, middle

# Two Continuous Variables

---

# What to look for

- Outliers
--

- Dependence or correlation
--

- Remember that correlation does not imply causation!
--

- Non linear relationships.

---

# Scatter plot

- For two metric variables use a scatter plot
  - One variable is represented by the `x` aesthetic
  - The other is represented by the `y` aesthetic
  - The geometry we use is `geom_point`.
- We will continue to use the `diamonds` dataset
  
---

# Scatterplot

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+geom_point()
  
```

---

# Overplotting

- When using big datasets, sometimes the points cover one another or are too close.
--

- This is sometimes called **overplotting**.
--

- Some solutions:
  - Try smaller points (size)
  - Try more transparent points (alpha)
  - Try a different geom
  
---

# Changing size

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_point(size=0.1)
  
```

---

# Changing alpha

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_point(alpha=0.2)
  
```
---

# Changing geom

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_bin2d()
  
```

---

# Hexagonal bins

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_hex()
  
```

---

# Changing geom

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_density2d()
  
```

---
# Time series plots

- When the x variable is time, it often makes more sense to join dots with a line.
--

- This way we can see 
  - Trend 
  - Seasonality 
  - Outliers
  - Structural break

---

# Economics dataset

- We will use the economics dataset (comes with ggplot2)

```{r, echo=TRUE}
str(economics)
```
- Notice date is its own type of variable

---

# Unemployed persons

```{r, echo=TRUE,fig.height=5}
ggplot(economics, aes(x=date, y=unemploy))+
  geom_line()

```

---
class:inverse, middle, center

# An aside on log scales
---
# Scale

- For variables that are heavily skewed it can be better to look at a log scale.
--

- For a regular scale you *add* as you move up the scale.
--

- For a log scale you *multiply* as you move up the scale.
--

- The log scale has the effect of putting more distance between smaller values and compressing higher values.

---


# Regular scale

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_point()
  
```

---


# Log scale


```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=carat,y=price))+
  geom_point()+scale_x_log10()+scale_y_log10()
  
```

---

# Zipf's Law

- In text mining, a well known empirical result is that the occurence of words in a document often follows Zipf's law

$$\mbox{Prob}(r)=\frac{r^{-s}}{K}$$

- Here $r$ is the rank of the word (1 is the most frequent, $N$ the least frequent).
- $K=\sum\limits^{N}_{x=1}x^{-s}$ is constant with respect to $r$.

---

# Three documents

- We will look at three documents:
  - The Australian Constitution
  - The script of Avengers Endgame
  - The homepage of online retailer Tao Bao.

---

# Australian Constitution

```{r, message=FALSE}
AusConsFreq<-read_csv('data/AusConstitution.csv')
AusConsFreq%>%
  filter(Rank<51)%>%
  ggplot(aes(x=reorder(Word,Rank),y=Freq))+
  geom_point()+
  labs(x='Word')+
  theme(axis.text.x = element_text(angle=90))
```

---

#Australian Constitution

```{r, message=FALSE}
AusConsFreq%>%
  ggplot(aes(x=Rank,y=Freq))+
  geom_point()
```

---

# Zipf Law

- Zipf's law predicts that

$$\mbox{Pr}(r)\approx r^{-s}/K$$

- Taking logs on both sides

$$log(f(r))\approx -slog(r)-log(K)$$

- Look at the plot on the log scale.

---

#Australian Constitution

```{r, message=FALSE}
AusConsFreq%>%
  ggplot(aes(x=Rank,y=Freq))+
  geom_point()+
  scale_x_log10()+
  scale_y_log10()
```

---
# Avengers Endgame

```{r, message=FALSE}
Avengers<-read_csv('data/Avengers.csv')
Avengers%>%
  filter(Rank<51)%>%
  ggplot(aes(x=reorder(Word,Rank),y=Freq))+
  geom_point()+
  labs(x='Word')+
  theme(axis.text.x = element_text(angle=90))
```

---

#Avengers Endgame

```{r, message=FALSE}
Avengers%>%
  ggplot(aes(x=Rank,y=Freq))+
  geom_point()
```

---

#Avengers Endgame

```{r, message=FALSE}
Avengers%>%
  ggplot(aes(x=Rank,y=Freq,label=Word))+
  geom_text(size=3)+
  scale_x_log10()+
  scale_y_log10()->g
ggplotly(g)%>%
  frameWidget(width = '100%',height = '100%')
```
---
# Tao Bao

```{r, message=FALSE}
TaoBao<-read_csv('data/TaoBao.csv')
TaoBao%>%
  filter(Rank<51)%>%
  ggplot(aes(x=reorder(Character,Rank),y=Freq))+
  geom_point()+
  labs(x='Character')+
  theme(axis.text.x = element_text(angle=90))
```

---

#Tao Bao

```{r, message=FALSE}
TaoBao%>%
  ggplot(aes(x=Rank,y=Freq))+
  geom_point()
```

---

#Tao Bao

```{r, message=FALSE}
TaoBao%>%
  ggplot(aes(x=Rank,y=Freq,label=Character))+
  geom_text(size=3)+
  scale_x_log10()+
  scale_y_log10()->g
ggplotly(g)%>%
  frameWidget(width = '100%',height = '100%')
```
---

# Other applications

- A similar observation is also made for the size of companies.
- Gibrat's Law claims that the growth rate of a company is independent of its size.
- This implies that the distribution of company size will be similar to the distribution of word frequency.
- Gibrat's law has also been applied to city populations.


---
class: inverse, middle, center

# Metric and Non-Metric Data

---

# Side by side plots

- When one variable is metric and the other non-metric we can easily put plots next to one another side by side.
- Simply map the non-metric variable to the x aesthetic and the metric variable to the y aesthetic.

---

# Boxplots

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=cut,y=price))+
  geom_boxplot()
  
```

---

# Change axes

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=price,y=cut))+
  geom_boxplot()
  
```

---

# With notches

- Recall that the notches provide a confidence interval around the median.
--

- These are particularly useful when comparing boxplots to one another.
--

- In general, if the confidence intervals overlap then the medians are not signficantly different.
--

- This is NOT a formal test, but still gives a useful indication.

---

# Boxplots (no overlap)

```{r, echo=TRUE,fig.height=5}
ggplot(data = mpg,
       mapping = aes(x=drv,y=hwy))+
  geom_boxplot(notch=T)
  
```

---
# Boxplots (some overlap)

```{r, echo=FALSE,fig.height=5}
filter(mpg,!(class%in%c('2seater','subcompact')))%>%
  ggplot(mapping = aes(x=class,y=hwy))+
  geom_boxplot(notch=T)
  
```

---
# Violin plot

- A violin plot is a newer visualisation.
- A kernel density is mirrored then arranged vertically.
- Specify the same way but use *geom_violin*
---

# Violin plot

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=cut,y=price))+
  geom_violin()
  
```

---

# Violin plot

```{r, echo=TRUE,fig.height=5}
ggplot(data = diamonds,
       mapping = aes(x=cut,y=price))+
  geom_violin()+coord_flip()
  
```

---

# Jittering

- A scatter plot can be used for non-metric data but can easily suffer from overplotting (one point on another). 

```{r, fig.height=5}
ggplot(data = mpg,
       mapping = aes(x=cyl,y=cty))+
  geom_point()

```

---

#Jittering

- Add random noise by jittering

```{r, echo=TRUE, fig.height=4}
ggplot(data = mpg,
       mapping = aes(x=cyl,y=cty))+
  geom_point(position = 'jitter')

```
