---
title: "Data Munging"
subtitle: "Data Visualisation and Analytics"
author: "Anastasios Panagiotelis"
date: "Lecture 6"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: extra.html
    
---

class: inverse, center, middle

# Data is messy

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(DT)
require(PoEdata)
require(kableExtra)
require(gridExtra)
require(ggmosaic)
require(GGally)
```

---

# Data 

- Data rarely comes in the fomat we want:
  - May not want to use all variables,
  - May not want to use all observations,
  - May need to transform variables.

---

# A few simple functions

- Much can be done with a few simple functions from the `dplyr` package:
  - Choose variables with `filter`
  - Choose observations with `select`
  - Transform variables with `mutate`
- In all cases both input and output is a data frame.

---

#Data frame

- A data frame is an object with:
  - Each row corresponding to an observation or case,
  - Each column corresponding to a variable.
- Two types of data frames in R are:
  - The `data.frame` in base R
  - The `tibble` in the tidyverse

---

# Diamonds

The diamonds data is a tibble

```{r, echo=TRUE}
diamonds
```

---

# Select variables

- The `select` function can be used if we only want to focus on a subset of variables.
- For the diamonds dataset we may only be interested in `carat`, `cut` and `price`.

```{r, echo=TRUE}
selected<-select(diamonds,carat, cut, price)
selected
```

---

# Select variables

- To drop variables, use a `-` sign

```{r, echo=TRUE}
select(diamonds,-carat, -cut, -price)
```

---

# Filter variables

- Suppose we only want to consider diamonds worth more than $1000
```{r,echo=TRUE}
filter(diamonds,(price>1000))
```

---

# Logical stamtents

- The term `(price>1000)` is an example of a logical statement.  It can be true or false.
- Other examples are
  - Price less than 1000 `(price<1000)`
  - Price less than or equal to 1000 `(price<=1000)`
  - Price not 1000 `(price!=1000)`
  - Price exactly 1000 `(price==1000)`
- Note **two** equals signs.
- In general `!` is the negation of a statement.

---

# And operator

- If two statements need to be satisfied use `&`
```{r,echo=TRUE}
filter(diamonds,
       (price>1000)&(cut=='Ideal'))

```

---

# Or operator

- If either one or the other statement needs to be satisfied use `|`
```{r,echo=TRUE}
filter(diamonds,(cut=='Ideal')|color=='E')
```

---

# In operator

Another useful operator is `%in%`
```{r,echo=TRUE}
filter(diamonds,
       (cut %in% c('Ideal','Fair')))
```

---

# Not operator

Use `!` as not 
```{r,echo=TRUE}
filter(diamonds,
       !(cut %in% c('Ideal','Fair')))
```

---
# Your turn

- Write R code for the following:
  - `price` greater than 2000 and `carat` less than 3.
  - `price` greater than 2000 and `cut` either Very Good, Premium or Ideal.
  - `carat` less than 2 or price greater than 500.
  - `carat` less than 2 and `cut` not equal to Premium.
---
class: inverse, center, middle
# Creating New Variables

---

# Mpg

Mpg is also a tibble

```{r, echo=TRUE}
mpg
```

---

# Mutate

- Suppose we are interested in fuel efficiency (the variables `cty` and `hwy`).
- These are measured in miles per gallon, but we want to convert into metric units.
- One mile per gallon equals 0.425144 km per litre
```{r,echo=TRUE}
mpgSel<-select(mpg,cty,hwy)
mpgMet<-mutate(mpgSel,ctyMetric=0.425144*cty,
                  hwyMetric=0.425144*hwy)
```
---

# Result

```{r,echo=FALSE}
mpgSel<-select(mpg,cty,hwy)
mpgMet<-mutate(mpgSel,ctyMetric=0.425144*cty,
                  hwyMetric=0.425144*hwy)
mpgMet
```
---

# The pipe operator

- The fact that the input and output are always data frames makes the pipe operator useful.
- Simply use `%>%` to chain commands together.
- Somtimes it is clearer to use `->` rather than use `<-` for assignment.
```{r, echo=TRUE}
c(1,12,4)%>%
  mean%>%
  sqrt->out
str(out)
```
---

#Pipes and dplyr

This code

```{r,echo=TRUE}
mpgSel<-select(mpg,cty,hwy)
mpgMet<-mutate(mpgSel,ctyMetric=0.425144*cty,
                  hwyMetric=0.425144*hwy)
```

does the same as this code

```{r,echo=TRUE}
select(mpg,cty,hwy)%>%
  mutate(ctyMetric=0.425144*cty,
                  hwyMetric=0.425144*hwy)->mpgMet
```

---
class: inverse, middle, center

# Grouping

---

# Group by function

- A powerful function in dplyr is the `group_by` function.
- Used in combination with `summarise`, it can construct  new datasets.
- Consider the `txhousing` dataset.  There is monthly data for each city.
- Suppose the aim is to find total number of sales for all cities in a year.

---

#Texas Housing

```{r, echo=TRUE}
txhousing
```
---

#Group by and summarise

```{r, echo=TRUE}
txhousing%>%
  group_by(year)%>%
  summarise(TotalSales=
              sum(sales,na.rm = TRUE))->TotSales
TotSales
```

---

#Group by two variables

- In earlier years, data for some cities is unavailable and is filled in as `NA`.
- The option `na.rm=TRUE` removes missing data before taking the sum.
- Next, suppose we want to get the yearly total for each city
- Need to group by `city` and `year`.


---
#Group by and summarise

```{r, echo=TRUE}
txhousing%>%
  group_by(year,city)%>%
  summarise(TotalSales=
              sum(sales,na.rm = TRUE))->TotSales
TotSales
```

---
#Your turn

- In the `diamonds` dataset, find the average price for each cut of diamond.
- In the `diamonds` dataset, find the average price for each cut of diamond given that price in above 2000.
- In the `mpg` dataset find the average fuel efficiency in the city of in each year.
- In the `mpg` dataset, consider Toyota, Nissan and Honda. Find the value of `hwy` for each of these manufacturer's most fuel efficient cars on the highway.



---

class: inverse, center middle

# About databases
---

# Codd's Model

- Many of the functions in `dplyr` are built on similar functions in the SQL language that is used to query databases.
- SQL is itself built upon (but not exactly the same) as a theoretical model known as Codd's model.
- An important aspect of this model is the first normal form.
- This also provide some guidelines for good data management.

---

# Example

Suppose we have the following database:

```{r}
eg<-tibble(Name=c('Ahmed','Bin','Carol'),
           DoB=c('1994/03/01','1954/12/23','1982/07/16'),
           Email=c('ahmed@personal.com',
                   'bin@me.com; bin@work.com',
                   'carol@mailcom'))  
kable(eg)%>%
  kable_styling(font_size = 24)
```

---

#Problems

- Carol's entry not a valid email.  
  - A function can be written to check and remove this.
- A bigger problem is that the email entry is not **atomic**.  There are two emails for Bin.
  - If code is written to check if the email is valid, then this code will also fail for Bin.
- On the next slide is a solution.

---

# Possible solution

```{r}
eg<-tibble(Name=c('Ahmed','Bin','Carol'),
           DoB=c('1994/03/01','1954/12/23','1982/07/16'),
           Email1=c('ahmed@personal.com',
                    'bin@me.com',
                    'carol@mailcom'),
           Email2=c(NA,
                    'bin@work.com',
                    NA))  
kable(eg)%>%
  kable_styling(font_size = 24)
```
---

#Problems

- This solution forces an ordering between Email1 and Email2 that may be arbitrary.
- Also, a new entry into the database may be: 
  - Name: Deepal
  - DoB: 1987/04/23
  - Email: deepal@work.com; deepal@me.com; coolgirl87@me.com
- The entire database needs to be changed to allow for three email addresses
---

# Solution

```{r}
eg<-tibble(Name=c('Ahmed','Bin','Bin','Carol'),
           DoB=c('1994/03/01','1954/12/23','1954/12/23','1982/07/16'),
           Email=c('ahmed@personal.com',
                    'bin@me.com',
                    'bin@work.com',
                    'carol@mailcom'))  
kable(eg)%>%
  kable_styling(font_size = 24)
```

Deepal can be added as

```{r}
eg<-tibble(Name=rep('Deepal',3),
           DoB=rep('1987/04/23',3),
           Email=c('deepal@work.com', 
                   'deepal@me.com', 
                   'coolgirl87@me.com'))  
kable(eg)%>%
  kable_styling(font_size = 24)
```

---

# First Normal Form

- This is only a very very superficial treatment of *database normalization*.
- The emphasis here is getting data into a form that is easily workable and minimises the chance for mistakes.
- It should be noted that the concept of *atomicity* is itself controversial and can be ambiguous.
- The main message is to think carefully about how you get data into an easily workable and robust form.

---

class: inverse, middle, center

# Reshaping

---

# Long v wide format

- Generally attempt to have one column per variable.
- Sometimes the meaning of a *variable* is ambiguous.
- In many cases it is easier to use `ggplot2` if the data are reshaped into a different form.
- We investigate using the `economics` data

---

# Economics wide

```{r, echo=TRUE}
economics
```

---

# Economics long

```{r, echo=TRUE, warning=FALSE}
economics_long
```

---

# Use wide

```{r,echo=TRUE,fig.height=5}
ggplot(economics,aes(x=uempmed,y=psavert))+
  geom_point()
```

---

# Use long

```{r,echo=TRUE,fig.height=4.5}
ggplot(economics_long,aes(x=date,y=value))+
  geom_line()+
  facet_wrap(~variable, scales = 'free_y')
```

---

# From wide to long

```{r,echo=TRUE}
gather(economics,key = Variable, 
       value = Value,-date)->long
long
```

---

# Gather function

- The `key` will be the name of the function that stores the variable name.
- The `value` will be the name of the variables that stores values.
- Use `-` to exclude variable(s) from gathering.
- Here this was done for date

---

#From long to wide

```{r,echo=TRUE}
economics_long%>%
  select(-value01)%>%
  spread(key = variable, 
         value = value)->wide
wide
```

---
class: inverse, center middle

# Data Import

---
# Data Import

- In many cases data is stored in a *flat database*.
- This means the data is stored in a single table.
- Columns are separated by a comma or a tab.
- Usually this table is contained in a single file.
- For reading in such data the `readr` package is very useful.

---

# The readr package

- Two of the more useful functions are `read_csv` and `read_tsv` for comma delimited and tab delimited files respectively.
- Generally the defaults for this function work quite well.
- One argument that is worth discussing a little is `col_types`

---

# Reading in long numbers

- The largest integer that many computers will accurately store is about 19 digits long.
- Some identification numbers are longer than that (for instance IBAN numbers used for bank transfers).
- In some cases, R may try to read these numbers in as integers.
- To avoid errors these should be read in as characters.

---

# Column types

- Specifying `col_types='c'` overrides the default behaviour of `readr`.
- You can use this to force everything to be read in as a string.
- If needed these can subsequently be converted to numeric variables using `as.numeric`.

---
# Summary

- With a small number of functions, much can be done
  - `select`, `filter`, `mutate`
  - `group_by`, `summarise`
  - `gather`, `spread`
- Mastering these functions takes practice.
- Often these functions are the key to creating plots.

