---
title: "Visualising One Variable"
subtitle: "Data Visualisation and Analytics"
author: "Anastasios Panagiotelis"
date: "Lecture 3"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: [default,"mtheme.css","mod.css"]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: extra.html
    
---

class: inverse, center, middle

# The Grammar of Graphics

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
require(magrittr)
require(tidyverse)
require(plotly)
require(widgetframe)
require(animation)
require(DT)
require(PoEdata)
require(kableExtra)
require(gridExtra)
require(ggmosaic)
```

---

# Visualising one variable

- Metric variables
  + Histogram
  + Density
  + Box plot
  + Rug plot
- Things to look for
  - Mode(s)
  - Skewness
  - Heavy tails
  - Outliers

---

# The grammar of graphics

- At first using `ggplot2` can seem too complicated.
- Once mastered it can be used to very easily create detailed plots.
- It is built on the ideas of *Grammar of Graphics* a text by Leland Wilkinson.
- The objective is to find an abstract set of rules for creating almost any graphic.

---

# Data

- The starting point for all visualisation is a dataset.
- Over the next few weeks, we will consider the datasets `diamonds`,  `mpg`, `economics` and `txhousing` which come built in with the `ggplot2` package.
- Later on we learn how to read in data.
- The `diamonds` data contains data on the price, size and quality of over 50000 diamonds.

---

# Aes and Geom

- Think of an *aesthetic* (or aes) as a way of perceiving a variable:
  + Position on x or y axis
  + Color
  + Size
- Think of a *geometry* (or geom) as a way of representing a variable:
  + Points
  + Lines
- `ggplot` maps aesthetics to geometries  

---
class: inverse, center, middle

#Histogram

---
# Histogram

- Consider a histogram of the variable price
- In a histogram, values of the variable we are interested in lie along the horizontal (x) axis.
- The histogram creates bins then counts the number of observations in each bin.
- To get started type

```{r, echo=T, eval=F}
ggplot(data = diamonds,mapping = aes(x=price))
```

---

# What do we see?

```{r, echo=F, eval=T}
ggplot(data = diamonds,mapping = aes(x=price))
```

---

# What do we see?

- We do have an x axis with a label price and some values.
- Otherwise we see nothing.
- We need to **add** a geometry to the plot.
- We do this with the `geom_histogram` function.
```{r, echo=T, eval=F}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram()
```

---

# What do we see?

```{r, echo=F, eval=T,message=FALSE}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram()
```

---

# Modification

- Suppose want to use a different number of bins or change the color of the bins?
- These are not features of the data or the aes
- These are features of the *geom*.  
- So these are controlled by arguments in the `geom_histogram` function.

---

# Change bins

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(bins = 5)
```

---

# Change boundary

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(bins =  5, boundary=0)
```
---

# Change binwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500)
```

---

# Change color

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,fill = 'red')
```

---

# Change border color

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,color='white',
                 fill = 'blue')
```

---
class:inverse, middle center

# An aside on colour

---

# Customise colour

- Many colours come built in to R.
- In some cases you may wish to select your own color.
- Customising colour requires appreciating how a computer understands color.
- We will do this by looking at RGB hex codes.
- Using this system, to a computer *#ff0000* is red.

---

# The RGB system

- One color model used by computers encodes every colour by the amount of *red*, *green* and *blue* light mixed to make that colour.
- This is called the RGB color model.
- A value between 0 and 255  indicates the strength of red, green and blue.
- These values between 0 and 255 are represented in two hexadecimal digits.

---

# Hexadecimal

- In hexadecimal: 
  - **a** is ten,
  - **b** is eleven,
  - **c** is twelve... 
  - **f** is fifteen.
- Take the first digit and multiply by 16 and add the second digit
- Hexadecimal is used since each digit corresponds to 4 bits in  computer memory.
---

#Examples

- **10** in hexadecimal is $1\times 16+0=16$ in decimal
- **1a** in hexadecimal is $1\times 16+10=26$ in decimal
- **2b** in hexadecimal is $2\times 16+11=42$ in decimal
- What is e4 in decimal?
---

# Color picker 

- One online tool to find the hex code of a color is <a href ="https://www.w3schools.com/colors/colors_picker.asp">here</a>.
- Suppose we want to the histogram to be <font color=#b35900> this brown</font> color.
- The hex code is #b35900 which is 179/256 red, 89/256 green and no blue.
- This can be provided as a string, to the `fill` or `color` argument of `geom_histogram`.
---

# Brown histogram

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(binwidth =  500,color='white',
                 fill = '#b35900')
```
---

# Finding hex codes

- It is useful to know hex codes since at times you may want to match colors for a specific purpose.
- For instance you may want the colors to match the brand colors of a client.
- For example a simple online search tells us that Coca Color red is <font color="#f40000">#f40000</font>
- The green color worn by NBA team the Milwaukee Bucks is <font color="#00471b">#00471b</font>.

---
# Histograms (Bucks Colors)

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_histogram(fill='#00471b',color='#eee1c6')
```
---

# An exercise

- Find the hex codes for a color(s) associated with:
  - A brand you like, or
  - A sports team you like, or
  - You country's flag,
  - Anything else
- Construct a histogram of the variable *carat* with these colors.

---
class: inverse, middle, center

# Density plot

---

# Density

- For a smoother version of a histogram we can use a different geom called `geom_density`.
- This in fact computes a kernel density estimate of the variable.
- The level of smoothness is controlled by a *bandwidth* parameter
- All the computation is done by `ggplot2`.

---

# Density plot

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density()
```

---
# Density plot (thicker)

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(size=3)
```

---
# Density plot (Bucks Colors)

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(fill='#EEE1C6',color='#00471B',size=3)
```

---
# How is density calculated?

- The kernel density estimate is a popular *nonparametric* technique that estimates a density as

$$\hat{f}(x)=\frac{1}{n}\sum\limits_{i=1}^{n}K_h(x-x_i)$$

- Here, $K_h()$ is a kernel function that depends on a bandwidth $h$.

---

# Uniform kernel

- The simplest kernel function is the uniform kernel
  - $K_h(u)=1/h$ if $|u|<h$
  - $K_h(u)=0$ otherwise
- At a point $x$, the estimated density is proportional to the number of points that are *close* to $x$.
- By *close*, we mean within $h$ units of $x$.

---
# Extremes

- If the bandwidth gets extremely large then for any $x$, all sample points are considered close.
- The formula for the kernel density becomes a flat line.
- If the bandwidth gets extremely small then for any $x$ we choose, the density is just the number of points in the sample equal to $x$.
- The kernel density is made up of spikes at the sample points.

---
# Defaults

- By default, `geom_density` 
  - Uses a Gaussian kernel 
  - Selects the bandwidth using *Silverman's rule of thumb*
- The same principles apply:
  - Large bandwidth leads to more smoothness
  - Small bandwidth leads to more bumpiness

---

# Density plot: Low bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=100)
```

---

# Density plot: High bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=2000)
```
---
# Density plot: Low bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=0.0001)
```

---

# Density plot: High bandwidth

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=price))+
  geom_density(bw=80000)
```
---

# Summary

- With both histograms and density plots
  - If the bin width or bandwidth is too small the plot may look bumpy.  This can exaggerate features that are not significant.
  - If the bin width or bandwidth is too large the plot may smooth over important features like local modes.
- Always try a few different values of bin width or bandwidth.

---
class: inverse, middle, center

# Finding outliers

---

# Outliers

- Histograms and density plots give a good idea of shape and local modes.
- Sometimes they can obscure outliers.
- For finding outliers a rug plot can be useful
- For finding outliers while still getting a good idea of skew, boxplots can be useful.
- We can investigate using the variable *carat*

---

# Carat: Histogram

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=carat))+
  geom_histogram()
```

---

# Carat: Rug plot

```{r, echo=T, eval=T,message=FALSE, fig.height=6}
ggplot(data = diamonds,mapping = aes(x=carat))+
  geom_rug()
```

---

# Box plot

- The box plot summarises 5 numbers
  - Median
  - First quartile $Q1$
  - Third quartile $Q3$
  - Upper Fence $U = Q3+1.5\times(Q3-Q1)$
  - Lower Fence $L = Q1-1.5\times(Q3-Q1)$
- Anything lying outside the fences represented as dots.
- When no points lie outside the fence, the fence is set to the maximum or minimum.

---

# Carat: Boxplot

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(y=carat))+
  geom_boxplot()
```

---

# Change of aesthetic

- Notice that the aesthetic changed!
- In the boxplot, the value of the variable is represented by the vertical (or y axis).
- We can change the definition of the upper and lower fence by passing the `coef` argument to `geom_boxplot`.  
- This changes the 1.5 used in calculating the fence to whatever you specify

---

# Changing fences

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(y=carat))+
  geom_boxplot(coef=4)
```


---

# Changing color

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = diamonds,mapping = aes(y=carat))+
  geom_boxplot(fill='#004d98',color='#a50044')
```

---

# Notches

- Notches can be added to a boxplot
- These are set to $\frac{1.58\times(Q3-Q1)}{\sqrt{n}}$
- This roughly gives a 95% confidence interval for the median.
- We will use a smaller dataset on the mileage of cars for this example to clearly illustrate the notches.

---

# Notches

```{r, echo=T, eval=T,message=FALSE, fig.height=5}
ggplot(data = mpg,mapping = aes(y=cty))+
  geom_boxplot(notch = T)
```

---
class: inverse, middle, center

# Non-metric data

---

# Nominal v Ordinal

- Non-metric variables are made up of **nominal** and **ordinal** variables.
- Nominal variables have no ordering in the categories of data:
  - Manufacturer of car (Audi, Toyota, etc).
- Ordinal variables do have an ordering in the categories:
  - Quality of diamonds (Fair, Good, etc).
  
---

# Non-metric variables in R

- Non-metric variables can be stored in R as
  - Character variables (nominal data)
  - Factors (nominal data)
  - Ordered factors (ordinal data)
- You can check with the `str` function

---

# Diamonds data

```{r, echo=T}
str(diamonds)
```

---

# Mpg data

```{r, echo=T}
str(mpg)
```

---

#Bar plot

- A common plot for non-metric data is the bar plot for the frequency of observations for each level of the factor.
- The height of each bar indicates the number of observations in a particular category.
- This can be done using `geom_bar`

---

#Bar plot

```{r, echo=T, fig.height=5}
ggplot(data = diamonds, mapping = aes(x=cut))+
  geom_bar()
```

---

#Bar plot

```{r, echo=T, fig.height=5}
ggplot(data = mpg, mapping = aes(x=manufacturer))+
  geom_bar()
```

---

# Summary

- Now you should be able to summarise a single variable.
- You should be able to change some features of the plot.
- You should also have some understanding of the grammar of graphics.
- Next we will look at summarising 2 (or more) variables on plots.